The role of priors, competence and relevance for the interpretation of disjunction \thanks{We thank \ldots}

### 2022-01-30 analysis/main/03_xor-some_prereg_analysis_final.Rmd
finish model comps, port over old intro

model_mean_posteriors_long %>%
  mutate(effect = factor(effect, 
                         levels = c("relevance_xor", "relevance_some", "prior_xor", "prior_some", "competence_xor", "competence_some"), 
                         labels = c("Relevance (or)", "Relevance (some)", "Prior (or)", "Prior (some)", "Competence (or)", "Competence (some)"))) %>%
  ggplot(.) + # 
  stat_halfeye(aes(y = effect, x = value, 
                   fill = stat(case_when(x %>% between(-0.05, 0.05) ~ "no effect", x < -0.05 ~ "negative effect", x > 0.05 ~ "positive effect"))
                   ), 
               alpha = .7) +
  geom_vline(xintercept = c(-.05, .05), linetype = "dashed") +
  scale_fill_manual(name = "Direction of effect", values = c("coral", "yellow3", "mediumseagreen")) +
  labs(x = "Effect size", y = "Effect") +
  theme(legend.position = "top",
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        axis.title = element_text(size = 13),
        strip.text.x = element_text(size = 13),
        strip.text.y = element_text(size = 13))

```

>>>>

model_mean_posteriors_long %>%
  mutate(effect = factor(effect, 
                         levels = c("relevance_xor", "relevance_some", "prior_xor", "prior_some", "competence_xor", "competence_some"), 
                         labels = c("Relevance (or)", "Relevance (some)", "Prior (or)", "Prior (some)", "Competence (or)", "Competence (some)"))) %>%
  ggplot(.) + # 
  stat_halfeye(aes(y = effect, x = value, 
                   fill = stat(case_when(x %>% between(-0.05, 0.05) ~ "no effect", x < -0.05 ~ "negative effect", x > 0.05 ~ "positive effect"))
                   ), 
               alpha = .7) +
  geom_vline(xintercept = c(-.05, .05), linetype = "dashed") +
  scale_fill_manual(name = "Direction of effect", values = c("coral", "yellow3", "mediumseagreen")) +
  labs(x = "Effect size", y = "Effect") +
  theme(legend.position = "top",
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        axis.title = element_text(size = 13),
        strip.text.x = element_text(size = 13),
        strip.text.y = element_text(size = 13))
#ggsave("../../writing/paper/images/posterior-effects-means.pdf", width=7, height=5)
```

### 2022-01-30 writing/paper/paper_draft.tex
finish model comps, port over old intro

General intuition, introduction \citep[e.g.,][]{horn1972semantic}.

@@

\subsection{Main analysis}
We analysed the results using a Bayesian linear mixed effects model, regressing the target implicature likelihood rating against the fixed effects of predictor ratings (i.e., the relevance, comptence and prior ratings elicited by the same participant for that vignette), the effect of trigger and their two-, three-, and four-way interactions. Maximal computationally tractable random effects structure justified by the experimental design was included: we included random intercepts and random slope effects for the main effects of trigger, relevance, competence and prior by-subject, as well as random intercepts by-vignette.\footnote{Model in R syntax style: \texttt{inference-rating $\sim$ trigger * relevance * competence * prior + (1 + trigger + relevance + competence + prior || subjectID) + (1 | vignette)}. For computational tractability reasons, the correlation of random effects by-subject was set to 0.} 
The categorical effect of trigger was dummy coded, using ``some'' as the reference level. For all regression coefficients we used a wide and uninformative prior given by a $t$-distribution with mean 0, standard deviation of 2 and 1 degree of freedom. The model was fitted using the R \texttt{brms} package \citep{burkner2017brms}.\footnote{The data and the analyses can be found under\newline \texttt{https://github.com/magpie-ea/magpie-xor-experiment}}

@@

\begin{figure*}[t]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-main.pdf}
	\end{center}
	\vspace{-0.3cm}
	\caption{Main analysis results: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect). Points indicate posterior means, thick lines indicate 66\% credible intervals, thin lines indicate 95\% credible intervals.}
	\label{posteriors-main}
\end{figure*}

@@

\subsection{Exploratory analysis}
\begin{figure*}[t]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-categorical.pdf}
	\end{center}
	\vspace{-0.3cm}
	\caption{Exploratory analysis with categorical predictors: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect). Points indicate posterior means, thick lines indicate 66\% credible intervals, thin lines indicate 95\% credible intervals. \mf{I'm not sure if we should blindly reuse a threshold of 0.05 here without further justification. Something to think about.}}
	\label{posteriors-cat}
\end{figure*}
In addition to the main analysis, we conducted two exploratory analyses. First, we fit a model using categorical predictors only. That is, we fit a Bayesian linear mixed effects regression model, regressing the inference rating against our prior intuitive classifications of the vignettes on the dimensions relevance, competence and prior (high or low, respectively), the effect of trigger, and their interactions. The random effects structure and the priors were identical to the main model. The factors were sum-coded, -1 representing the low condition, 1 representing the high condition. 
This model decouples the predictors by dichotomizing their classification and, therefore, addresses the potential collinearty issue. Furthermore, it provides a comparison between participants' judgements and our intuitive prior classifications of the vignettes.

Given this model and the data, we found very similar results for both triggers: We observe a negative effect of prior and a positive effect of competence, in line with predictions of the scalar implicature hypothesis (Fig.\ref{posteriors-cat}). However, we again observe no effects of relevance. If anything, the relevance effect tends in the direction opposite to predictions of the SI account for the trigger ``or''. 
\begin{figure*}[t]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-means.pdf}
	\end{center}
	\vspace{-0.3cm}
	\caption{Exploratory analysis with predictor ratings averaged by-vignette: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect). Points indicate posterior means, thick lines indicate 66\% credible intervals, thin lines indicate 95\% credible intervals. \mf{I'm not sure if we should blindly reuse a threshold of 0.05 here without further justification. Something to think about.}}
	\label{posteriors-mean}
\end{figure*}

@@

>>>>

General intuition, introduction \citep[e.g.,][]{horn1972semantic}.
 \pt{ported over from old paper below}
Introductions to logic usually distinguish between two readings of “or”: an inclusive
and an exclusive reading (e.g., McCawley, 1981; Copi \& Cohen, 2005). The inclusive
reading corresponds to the meaning of logical disjunction. According to this reading,
“A or B” is true if at least one and possibly both of A and B are true. The exclusive
reading is more strict in that it excludes the possibility that both A and B are true. So
on its exclusive reading, “A or B” is true if exactly one of A and B is true. The presence of these two readings might suggest that “or” is associated with two
different lexical entries (e.g., Basson \& O’Connor, 1960; Baum, 1996; Rescher, 1964).
There are, however, good reasons to reject such a lexicalist approach. Perhaps the most
compelling reason is that, in some contexts, “or” can only receive one interpretation.
To illustrate, consider the following sentence, in which “or” occurs in the scope of
negation: (2) Joe does not support Donald or Hillary.
This sentence has only one interpretation, namely that Joe supports neither Donald nor
Hillary (cf. Crain, 2008). This interpretation corresponds to the inclusive reading of
“or”. It does not have an interpretation corresponding to the exclusive reading; in other
words, it does not have a reading according to which Joe either supports both Donald
and Hillary, or neither of them. Since “or” is systematically monosemous in certain
contexts—negation being one of them—it follows that the two readings of “or” cannot
simply be due to a lexical ambiguity.
An attractive alternative to the lexicalist approach stems from Grice’s (1975) theory
of conversational implicature. According to Grice, natural conversation is governed by
the assumption that speakers are cooperative; that is, they attempt to further the purpose
of the discourse by means of their utterances. Speakers are cooperative by adhering
to four maxims that enjoin their utterances to be truthful, informative, relevant,
and clear. Sometimes hearers have to make ancillary assumptions in order to align a
speaker’s utterance with the assumption of cooperativity. Such ancillary assumptions
are called conversational implicatures. Horn (1972) was the first to argue that the exclusive reading of “or” can be explained
as a conversational implicature, too, based on the assumption that the primary
meaning of “or” is inclusive. To illustrate, consider (1) again. Assuming that the primary
meaning of “or” is inclusive, the speaker of (1) could have been more informative,
and hence cooperative, by uttering the “and”-alternative “Joe supports Donald
and Hillary.” Why didn’t she? Presumably because she does not believe that the “and”-
alternative is true. This is a weak inference that is compatible with a situation in which
the speaker is unsure about whether or not Joe supports both Donald and Hillary. This
weak inference can be strengthened if there is reason to believe that the speaker knows
whether or not Joe supports both Donald and Hillary. This assumption is often called
the competence assumption (e.g., Geurts, 2010; Russell, 2006; Schulz \& van Rooij,
2006; Zimmermann, 2000). If the competence assumption is sufficiently plausible, it
follows that, according to the speaker, Joe does not support both Donald and Hillary.
This specific kind of conversational implicature is often called a scalar implicature
because it is assumed that “or” forms a lexical scale with “and”, and that alternatives
are generated by substituting the scalemates “or” and “and”. Other lexical scales are
hsome, alli, hwarm, hoti, and hintelligent, brillianti.

The implicature account straightforwardly explains the absence of exclusive readings
under negation. Consider (2) again. Here, unlike in the case of (1), the “and”-
alternative is less informative than the utterance itself. After all, (2) is only compatible
with a situation in which Joe supports neither Donald nor Hillary, whereas the
“and”-alternative is also consistent with situations in which Joe supports exactly one
of Donald and Hillary. So the speaker was already maximally informative and hence
no implicature is derived.
Although the implicature account has since become the standard in the literature
(e.g., Chevallier, Noveck, Nazir, Bott, Lanzetti, \& Sperber, 2008; Chierchia, Fox, \&
Spector, 2012; Geurts, 2010; Sauerland, 2004), it is not without its problems, as we
will see in the next section. Therefore, we set out to experimentally test the adequacy
of the implicature account. To that end, we investigated the effect of three factors on the
robustness of exclusive readings: (i) the relevance of the “and”-alternative to the hearer,
(ii) the competence of the speaker about the truth of the “and”-alternative, and (iii) the
prior probability that the “and”-alternative is true. These factors are usually taken to
influence the robustness of scalar implicatures. Hence, if the implicature account is
correct, we expect these factors to influence the robustness of the exclusive reading of
“or”, too, For comparison, we also investigated how relevance, competence, and prior probability
influence the robustness of a bona fide scalar implicature, namely the inference
from “some” to “not all”. To illustrate, consider:
(4) Some of my friends support Donald.
An utterance of this sentence may convey that not all of the speaker’s friends support
Donald. The derivation of this upper-bounded construal runs analogous to that of the
exclusive reading of “or”: the speaker could have been more informative by saying
“All of my friends support Donald”. Why didn’t she? Presumably because she does. not believe that this alternative is true. If, moreover, the speaker knows whether or
not all of her friends support Donald—i.e., if the speaker is competent in the relevant
sense—it follows that she believes that not all of her friends support Donald.
As we will see, it turns out that there are marked differences in how relevance,
competence, and prior probability influence the robustness of the exclusive reading of
“or” compared to how they influence the robustness of the inference from “some” to
“not all”. These findings will encourage us to reconsider the implicature account of the
exclusive reading of “or”. In the next section, we explain our three factors of interest in more detail. Afterwards, we outline the details of our experiments and discuss the results.

@@

\subsection{Predictor ratings}

\begin{figure}[h]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/byFactor-byPrior-raw-scatter.pdf}
	\end{center}
	\vspace{-0.3cm}
	\caption{Relating ratings for relevance, competence and prior statements (x-axis) to ratings for the strength of pragmatic enrichments (y-axis). The top row shows ratings for “some” (enriched to “some, but not all”). The bottom row shows ratings for “or” (enriched to “A or
B, but not both”). Ratings for stories categorized as low (red) w.r.t. a given factor are on average lower (x-axis) than for those categorized as high (blue). The apparent effect of prior for
“or” is mitigated by collinearity with relevance.}
	\label{main-raw-ratings}
\end{figure}

Participants’ factor ratings by-story agreed well with the designed classification of the stories (Fig. \ref{main-raw-ratings}, red vs. blue color on x-axis).

\subsection{Main analysis}

We analysed the results using a Bayesian linear mixed effects model, regressing the target implicature likelihood rating against the fixed effects of predictor ratings (i.e., the relevance, comptence and prior ratings elicited by the same participant for that vignette), the effect of trigger and their two-, three-, and four-way interactions. Maximal computationally tractable random effects structure justified by the experimental design was included: we included random intercepts and random slope effects for the main effects of trigger, relevance, competence and prior by-subject, as well as random intercepts by-vignette.\footnote{Model in R syntax style: \texttt{inference-rating $\sim$ trigger * relevance * competence * prior + (1 + trigger + relevance + competence + prior || subjectID) + (1 | vignette)}. For computational tractability reasons, the correlation of random effects by-subject was set to 0.} 
The categorical effect of trigger was dummy coded, using ``some'' as the reference level. For all regression coefficients we used a wide and uninformative prior given by a $t$-distribution with mean 0, standard deviation of 2 and 1 degree of freedom. The model was fitted using the R \texttt{brms} package \citep{burkner2017brms}.\footnote{The data and the analyses can be found under\newline \texttt{https://github.com/magpie-ea/magpie-xor-experiment}}

@@

\begin{figure}[h]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-main.pdf}
	\end{center}
	\vspace{-0.3cm}
	\caption{Main analysis results: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect). Points indicate posterior means, thick lines indicate 66\% credible intervals, thin lines indicate 95\% credible intervals.}
	\label{posteriors-main}
\end{figure}

@@

\subsection{Exploratory analysis}
\subsubsection{Models with alternative predictors}
\begin{figure}[h]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-categorical.pdf}
	\end{center}
	\vspace{-0.3cm}
	\caption{Exploratory analysis with categorical predictors: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect). Points indicate posterior means, thick lines indicate 66\% credible intervals, thin lines indicate 95\% credible intervals. \mf{I'm not sure if we should blindly reuse a threshold of 0.05 here without further justification. Something to think about.}}
	\label{posteriors-cat}
\end{figure}

In addition to the main analysis, we conducted three exploratory regression analyses. First, we fit a model using categorical predictors only. That is, we fit a Bayesian linear mixed effects regression model, regressing the inference rating against our prior intuitive classifications of the vignettes on the dimensions relevance, competence and prior (high or low, respectively), the effect of trigger, and their interactions. The random effects structure and the priors were identical to the main model. The factors were sum-coded, -1 representing the low condition, 1 representing the high condition. 
This model decouples the predictors by dichotomizing their classification and, therefore, addresses the potential collinearty issue. Furthermore, it provides a comparison between participants' judgements and our intuitive prior classifications of the vignettes.

Given this model and the data, we found very similar results for both triggers: We observe a negative effect of prior and a positive effect of competence, in line with predictions of the scalar implicature hypothesis (Fig.\ref{posteriors-cat}). However, we again observe no effects of relevance. If anything, the relevance effect tends in the direction opposite to predictions of the SI account for the trigger ``or''.

\begin{figure}[h]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-means.pdf}
	\end{center}
	\vspace{-0.3cm}
	\caption{Exploratory analysis with predictor ratings averaged by-vignette: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect). Points indicate posterior means, thick lines indicate 66\% credible intervals, thin lines indicate 95\% credible intervals. \mf{I'm not sure if we should blindly reuse a threshold of 0.05 here without further justification. Something to think about.}}
	\label{posteriors-mean}
\end{figure}

@@

\subsubsection{Model comparisons}
Because the correlations hint at possible collinearity effects between relevance and prior for the trigger ``or'', we conduct exploratory model comparisons. That is, in order to check whether certain factors among relevance, competence and prior have a stronger influence on the inference stength ratings, we compare regression models of different complexity. We take a Bayesian approach to model comparison here in terms of their Bayes factors \citep{rouder2012default}. 
We compare all regression models that can be built with the three explanatory factors as main factors against a baseline intercept-only model.\footnote{If the model contains more than one main effect, all interactions between main effects are always included in the model, as well. All justified and computationally tractable random effects are added to each model, respectively.}
Following literature, we interpret the evidence in favor of a particular model as strong if the odds (i.e., the Bayes Factor) are between are above 10, as mild, if they are between 3 and 10, and as absent if the odds are between 3 and 1/3 \citep[e.g.,][]{lodewyckx2011tutorial}.

\begin{figure}[h]
	\begin{center}
		\includegraphics[width=0.8\linewidth]{images/BF_vs_SD_allModels_vs_int.pdf}
	\end{center}
	\vspace{-0.3cm}
	\caption{Bayes factor comparison of different main factor combinations.}
	\label{bf-grid-search}
\end{figure}

To investigate the stability of the Bayes factors of a certain model over the baseline, we conduct a grid search over the standard deviation parameter of the prior student-t distribution over the main effect regression coefficients. More precisely, given that, based on prior experimental work, the expected effect size for the main factors was around $\beta=0.15$, the standard deviation parameters considered in the grid search included {0.05, 0.1, 0.15}. Figure \ref{bf-grid-search} shows the Bayes factors against the intercept-only model of all the regression models plotted against the different standard deviation parameters of the priors. It indicates that, in general, the Bayes factor does not change strongly dependent on the tightness of the prior distribution. It further suggests that the intercept-only model is a better fit to the data than all of the explored regression models, but more so for models containing both relevance and prior as explanatory factors. This corroborates the finding based on the correlation, indicating that relevance and prior might be co-dependent and explain each other away. We also observe that models containing the effect of relevance are slightly less likely given the data than models containing the effect of prior, but the difference is negligible. However, these results might also be due to the fact that additional model complexity of the regression models overpowers potential improvement in model fit.

\begin{figure}[h]
	\begin{center}
		\includegraphics[width=0.8\linewidth]{images/BF_vs_SD_rel-comp_pri-comp.pdf}
	\end{center}
	\vspace{-0.3cm}
	\caption{Bayes factor of the relevance*competence model over the prior*competence model.}
	\label{bf-rel-pri}
\end{figure}

Therefore, to zoom in on the comparison between the effects of relevance and prior on inference strength ratings we conduct further exploratory model comparisons. Specifically, we compare two-factor models, comparing relevance over prior, each combined with competence, respectively. As can be seen in Figure \ref{bf-rel-pri}, the Bayes factor varies more strongly depending on the standard deviation parameter of the prior distribution over the main regression coefficients. For $SD=0.05$, we find $BF=0.27$, indicating that ``prior'' model is around three times more likely than the ``relevance'' model, which can we consider mild evidence in favor of prior being the better predictor for the inference strength of exclusive disjunctions. 
Yet further investigation of the interaction between relavance and competence effects with respect to their influence on exclusive disjunctions is needed.
